---
title: "Happiness Project"
author: "Sujata Verma"
date: "8/11/2020"
output:
  word_document: 
    fig_caption: yes
    fig_height: 6
    fig_width: 10
    keep_md: yes
    toc: yes
  pdf_document: default
---
#                             INTRODUCTION

I will be exploring dataset on World Happiness Report 2020 which is based on the global survey of 156 countries.  Data is collected through surveys from 2017-19 conducted by Gallup World Poll. The survey asked repondants to rate their life on multiple parameters on a scale from 0 to 10, with 10 being the best and 0 being the worst. This ranking is called the Cantril Ladder. The six determinants are GDP per person, healthy life expectancy, social support, perceived freedom to make choices about one's life, generosity, and perception of corruption. The report defines Dystopia as an imaginary country with the worst possible score in all variables. The dystopia plus residual factor captures the three-year average lowest score of 1.97 and the unexplained variation from each of the six determinants.

In my project, I will add another explanatory variable, income ineqaulity within each country, as measured by the Gini Coefficient.

Per-capita GDP captures the economic well being of the residents of a country but the GDP is not equitably distributed within the country. My hypothesis is that when people are asked about their economic well being, the disparities in income influences their perception about their well-being and hence their happiness score. 

In order to find out the affect of income inequality in a country on its happiness score, I will combine the World Happiness Report data set with Gini coefficients data from the World Bank for various countries, for the latest year that is available. I will then check for missing values and try and update the data set. Gini coefficient information from the CIA World Factbook will be used to supplement the data for countries where World Bank data is not available.  The Gini coefficents will be scaled from 0 (perfect equality) to 1 (perfect inequality).

#                     GOAL OF THE PROJECT AND MAIN STEPS

The goal of this project is to explore what determines happiness of people. More specifically, does income inequality within nations affect the Happiness Scores? I will also try and fit the best predictive model so that it accurately describes the determinants of happiness.

Main steps include creating the combined dataset, data visualization, and fitting models with the best predictive power.

Creating the dataset:

First step is to load the libraries and options that will be needed.


```{r echo=FALSE, message=FALSE, warning=FALSE}


#load libraries
library(tidyverse)
library(ggplot2)
library(ggthemes)
library(caret)
library(ggrepel)
library(corrplot)
library(ggpubr)
library(rpart)
library(rpart.plot)
library(randomForest)
library(tinytex)
if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
if(!require(caret)) install.packages("caret", repos = "http://cran.us.r-project.org")
if(!require(data.table)) install.packages("data.table", repos = "http://cran.us.r-project.org")
if(!require(corrplot)) install.packages("corrplot", repos = "http://cran.us.r-project.org")
if(!require(rpart)) install.packages("rpart", repos = "http://cran.us.r-project.org")
if(!require(rpart.plot)) install.packages("rpart.plot", repos = "http://cran.us.r-project.org")
if(!require(randomForest)) install.packages("randomForest", repos = "http://cran.us.r-project.org")
if(!require(ggpubr)) install.packages("ggpubr", repos = "http://cran.us.r-project.org")
if(!require(ggplot2)) install.packages("ggplot2", repos = "http://cran.us.r-project.org")
if(!require(tinytex)) install.packages("tinytex", repos = "http://cran.us.r-project.org")
if(!require(readr)) install.packages("readr", repos = "http://cran.us.r-project.org")
if(!require(readxl)) install.packages("readxl", repos = "http://cran.us.r-project.org")

options(digits=3)

```
Second step is the load the two data-sets,  from the World Happiness Report 2020 and the World Bank website and then to combine them into a single data set. The two data files are available on my github page.

Datasets:

https://raw.githubusercontent.com/econverma/The-Happiness-Project/master/WHR20.csv

https://raw.githubusercontent.com/econverma/The-Happiness-Project/master/gini%20updated.csv


Here is the structure of the variables from the Happiness Report dataset:

```{r echo=FALSE, message=FALSE, warning=FALSE}
#load happiness report data
file<-read_csv(url("https://raw.githubusercontent.com/econverma/The-Happiness-Project/master/WHR20.csv"))


#select columns
data<-file%>%
  select(c(`Country name`,`Regional indicator`,`Ladder score`,`Logged GDP per capita`,`Social support`,`Healthy life expectancy`,`Freedom to make life choices`,Generosity,`Perceptions of corruption`,`Dystopia + residual`))

#rename columns 

colnames(data)<-c("country","region","score","log_pcgdp","soc_support","life_exp","freedom","generosity","corruption","dys_res")

#explore data set
class(data)
str(data)      


#changing country and region from charaters to factors

data$country<-as.factor(data$country)
data$region<-as.factor(data$region)
#country variable where the surveys were conducted
#regions
#score is the overall happiness score from 0 to 10
#log_pcgdp are the World Bank per-capita GDP in constant 2011 dollars in Purchasing parity parity terms
#soc_support is national average to the question"If you were in trouble, can you count on friends/family to help". Ansswer is binary 0 or 1.
#life_exp is based on WHO data on life expectancy at birth
#freedom is national average to the question"Are you satisfied or dissatisfied with the freedom to choose what you do in your life" Answers are binary.
#generosity is the residual of regressing national average of responses to "have you donated to a charity in the past month?" to GDP per capita.
#corruption is average of responses to "Is their widespread corruption in Government/Business?"
#dys_res is the value of 1.97 as the lowest benchmark value + any residual
```




Next, I imported the income inequality data from World Bank and combine the two datasets, adding gini coefficient for countries  as an explanatory variable. The World Bank data file was updated for missing values with additional information from CIA World Factbook data, obtained from the Wikipaedia page. Also, names of some of the countries were updated so that they were written exactly the same way as the World Happiness report dataset. The column names were also simplified. The Gini coefficents were scaled from 0 (perfect equality) to 1 (perfect inequality). Here are the all the variables contained in the final data set.

Country variable includes countries where the surveys were conducted.

Regions variable defines the geographical regions where a country is located.

Score is the overall happiness score from 0 (least happy) to 10 (most happy).

Log_pcgdp are the World Bank estimates of per-capita Gross Domestic Product in constant 2011 dollars in Purchasing Power Parity terms.

Soc_support is national average to the question"If you were in trouble, can you count on friends/family to help?". Answer is binary and takes the values of 0 or 1.

Life_exp is based on WHO data on life expectancy at birth.

Freedom is national average to the question"Are you satisfied or dissatisfied with the freedom to choose what you do in your life" Answers are binary.

Generosity is the residual of regressing national average of responses to "Have you donated to a charity in the past month?" to GDP per capita.

Corruption is average of responses to "Is there widespread corruption in Government/Business?"

Dys_res is the value of 1.97 as the lowest benchmark value + any residual

Gini Coefficient captures income inequality within a country from 0(most equality) to 10(most inequality).

Once we add the data on Gini Coeffiecients, we explore the summary statistics for each of the variables. It will also let us know if there are any missing values in the data set. We can then remove the missing values and check the dimensions of the final data set.

```{r,echo=FALSE , message=FALSE, warning=FALSE}


#load the income inequailty data from World Bank and combine the two data sets

gini_updated<-read_csv(url("https://raw.githubusercontent.com/econverma/The-Happiness-Project/master/gini%20updated.csv"))
ginidata<-data%>%left_join(gini_updated,by="country")%>%as.data.frame()

ginidata<-ginidata%>%mutate(gini=Gini/100)

ginidata<-ginidata%>%select("country","region","score","log_pcgdp","soc_support","life_exp","freedom","generosity","corruption","dys_res","gini")

#examine the data and remove missing values
summary(ginidata)

ginidata<-ginidata%>%remove_missing()

#changing country and region from charaters to factors

ginidata$country<-as.factor(ginidata$country)
ginidata$region<-as.factor(ginidata$region)


dim(ginidata)

```

This completes the first step of importing and combining datasets and removing the missing values. We now have data on 138 countries and 11 variables with no missing values. The variables include: country name, region, happiness score,log of per-capita GDP, social support, life expectancy, freedom, generosity, corruption, dystopia (or minimum score)+ residuals, and gini coefficent. The dependent variable is the happiness score.


#                            DATA VISUALIZATION


Let us take a look at the correlation among the variables by building a correlation plot. It will tell us which are the important explanatory variables affecting the happiness score and whether the explanatory variables have a positive or negative effect on the happiness score. We exclude country name and region for the correlation plot. Note that correlation doesn't imply cause and effect.

```{r echo=FALSE, message=FALSE, warning=FALSE}

# Create a correlation plot among all the variables
cor = cor(ginidata[c(3:11)], method = "spearman")
corrplot::corrplot(cor,method="number")

```


From the correlation plot, we can infer that the Growing Economy, Social Support and Life Expectancy have the biggest positive correlation with the Happiness Score. Freedom to make choices also correlates with Happiness positively. Corruption and Income Inequality have a weak negative correlation with Happiness, while Generosity is not correlated with the Happiness Score. The Dystopia + Residual is positively correlated to Happiness, showing that we are not capturing all of the determinants of happiness.

Other interesting strongly positive correlations are between the Growing Economy and Life Expectancy and also between the Growing Economy and Social Support. Gini coefficient has a weak negative correlation with Per-capita GDP, Social Support as well as Life Expectancy. 

These relationships among Happiness Score and each of the seven explanatory variables is further explored by a country-wise scatter plot for each of the variables against the dependent variable, Happiness Score. Instead of a linear regression line, Loess method is used to get a smooth curve.

Note,we don't explore Dystopia + Residuals in the future analysis because it is not a factor causing happiness or unhappiness and we focus on what makes people happy.

Economy and Happiness
```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data, aes(x=log_pcgdp, y=score, label=country))+
  geom_point()+
  geom_smooth()+ylab('Happiness Score')+xlab('Per-capita GDP')+
  geom_text(aes(color=country),size=2)+theme(legend.position = "none")

```

The above graph confirms that wealthier the country, more is the reported happiness.

Life Expectancy and Happiness
```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data, aes(x=life_exp, y=score, label=country))+
  geom_point()+
  geom_smooth()+ylab('Happiness Score')+xlab('Life Expectancy')+
  geom_text(aes(color=country),size=2)+theme(legend.position = "none")
```

The healthier the population of a country as denoted by life-expectancy, more is the reported happiness. The curve levels off at very low and very high life-expectancy rates.

Social Support and Happiness
```{r echo=FALSE, message=FALSE, warning=FALSE}
#Family and friends support and happiness
ggplot(data, aes(x=soc_support, y=score, label=country))+
  geom_point()+
  geom_smooth()+ylab('Happiness Score')+xlab('Social Support')+
  geom_text(aes(color=country),size=2)+theme(legend.position = "none")
  

```

The relationship between friends and family support and happiness is a very strong one. The happiness increases exponentially as the family support increases. Western European countries and Australia and New Zealand ranked at the top in terms of social support. Interestingly, Turkmenistan and Slovenia ranked in the top 10 in terms of family and social support.

Freedom and Happiness
```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data, aes(x=freedom, y=score, label=country))+
  geom_point()+
  geom_smooth()+ylab('Happiness Score')+xlab('Freedom')+
  geom_text(aes(color=country),size=2)+theme(legend.position = "none")
  
```

The chart above shows that happiness is higher in countries where residents believe that they have the freedom to make decisions about their life.

Generosity and Happiness
```{r echo=FALSE, message=FALSE, warning=FALSE}
#Generosity and Happiness
ggplot(data, aes(x=generosity, y=score, label=country))+
  geom_point()+
  geom_smooth()+ylab('Happiness Score')+xlab('Generosity')+
  geom_text(aes(color=country),size=2)+theme(legend.position = "none")
```

Being charitable and generous doesn't make us happier, according to the above chart. The three most generous countries were Myanmar, Indonesia and Haiti, yet the people in these countries are not among the richest or the happiest.

Corruption and happiness
```{r echo=FALSE, message=FALSE, warning=FALSE}
#Happiness and perception of corruption
ggplot(data, aes(x=corruption, y=score, label=country))+
  geom_point()+
  geom_smooth()+ylab('Happiness Score')+xlab('Corruption')+
  geom_text(aes(color=country),size=2)+theme(legend.position = "none")

```

There is a negative relationship between perceptions of corruption and happiness. What is interesting about this chart is the preponderance of multiple counties at a very high level of corruption. It would seem that corruption is a way of life in most of the world.

Income Inequality and Happiness
```{r echo=FALSE, message=FALSE, warning=FALSE}
#Happiness and Income Inequality
ggplot(ginidata, aes(x=gini, y=score, label=country))+
  geom_point()+
  geom_smooth()+ylab('Happiness Score')+xlab('Income Inequality')+
  geom_text(aes(color=country),size=2)+theme(legend.position = "none")
```

There is a negative relationship between Income Inequality and Happiness. This chart supports the hypothesis that income disparities lowers happiness.

            Regional Income Inequalities and Happiness:

In this section, we will explore the link between happiness and income inequality by region. Let us first take a look at the previous chart, but this time dis-aggregating by region. We abbraviate the ten regions as follows:
Central and Eastern Europe(CaEE),	Commonwealth of Independent States(CoIS)	, East Asia(EstA), Latin America and Caribbean(LAaC), Middle East and North Africa(MEaNA), North America and ANZ(NAaA), Sub-Saharan Africa (S-SA), South Asia(SthAs), Southeast Asia (SthsA), and Western Europe (WstE).


```{r echo=FALSE, message=FALSE, warning=FALSE}
#Scatter-plots of Income Inequality by region and countries


ggplot(ginidata, aes(x=gini, y=score))+
  geom_point()+ylab('Happiness Score')+xlab('Income Inequality')+
  geom_text(aes(color=country,label=country),size=2)+theme(legend.position = "none")+ facet_grid(~abbreviate(region))
  
  
```

We can see from the above chart, that Sub-saharan Africa and Latin America and the Carribean have a higher Inequality and a lower Happiness Score. While North America, Australia and New Zealand as well as Western Europe have a lower Income Inequality and higher Happiness Score.

Instead of scatter plots, we can have a clearer picture of the relationship between region-wise Income Ineqalities and Happiness by creating box-plots.

        Box-plots of Regional Income Inequality and Happiness
```{r echo=FALSE, message=FALSE, warning=FALSE}
# aggregate by region and create box-plots by income inequality and happiness scores

#By happiness scores
g1<-ggplot(ginidata, aes(x=region, y=score,fill=region))+
  geom_boxplot()+ylab('Happiness Score')+xlab('Region')+
  theme(axis.text.x = element_text(angle = 60, hjust = 1))+
  theme(legend.position = "none")

#By income inequality

g2<-ggplot(ginidata, aes(x=region, y=gini,fill=region))+
  geom_boxplot()+ylab('Income Inequality')+xlab('Region')+
  theme(axis.text.x = element_text(angle = 60, hjust = 1))+
  theme(legend.position = "none")

#arrange the box-plots side-by-side
ggarrange(g1,g2,ncol=2)




```

From the figure on the left, we can infer that Western Europe and North America, Austraia and NewZealand are the happiest regions. Sub-Saharan Africa, Middle East and North Africa and South Asia score less on the Happiness Score.

From the figure on the right, Latin America and Caribbean and Sub-Saharan Africa are some of the most inequitable regions and Western Europe is the one of the most equitable regions.

Latin America and Caribbean is an exception to this inverse relationship as it has high Happiness Score despite having a high degree of Income Inequality.



#                            METHOD AND ANALYSIS


The goal of the model is to see how well different explanatory variables predict the happiness of a country. We will apply different models to the training and testing data. The best model will be the one with the lowest Root Mean Square Error (RMSE), which calculates how far is the actual happiness score from the score predicted by the model.

First step is to create training and testing data sets. I decided to split the entire data set containing 138 rows equally between the training and testing data-sets, since the number of observations is not very large, so an uneven split will make the results for the smaller data set very skewed.

```{r echo=FALSE, message=FALSE, warning=FALSE}
##create training and testing set with p=0.5


#remove country,region and dys-res from the data
hdata<-ginidata%>%select("score","log_pcgdp","soc_support","life_exp","freedom","generosity","corruption","gini")


set.seed(1, sample.kind="Rounding")


#creat training and testing set with p=0.5 since the data set is small.

test_index <- createDataPartition(y = hdata$score, times = 1, p = 0.5, list = FALSE)
train_set <- hdata[-test_index,]
test_set <- hdata[test_index,]



```

# Which Regression Model will be appropriate? 

We have one dependent variable, happiness score, and seven explanatory variables. But do these variables follow a normal distribution? To find out we create smooth density functions of each of these variables below.

```{r echo=FALSE, message=FALSE, warning=FALSE}
#check for multi-variate normal distribution
h1<-ggplot(ginidata,aes(x=score))+geom_density()
h2<-ggplot(ginidata,aes(x=log_pcgdp))+geom_density()
h3<-ggplot(ginidata,aes(x=soc_support))+geom_density()
h4<-ggplot(ginidata,aes(x=life_exp))+geom_density()
h5<-ggplot(ginidata,aes(x=freedom))+geom_density()
h6<-ggplot(ginidata,aes(x=generosity))+geom_density()
h7<-ggplot(ginidata,aes(x=corruption))+geom_density()
h8<-ggplot(ginidata,aes(x=gini))+geom_density()

ggarrange(h1,h2,h3,h4,h5,h6,h7,h8,nrow=4,ncol=2)
```

As we can see from the above chart, the variables don't show a normal distribution. Instead of a multivariate normal model, it will be more appripriate to apply a generalized linear regression model (GLM).

##  1. GLM Model

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Fitting Generalized Linear Model Regression to the training set

glm_method = glm(formula = score ~ .,
                  data = train_set)

summary(glm_method)


#Predicting the Test set results
p_hat_glm = predict(glm_method, newdata = test_set)


RMSE_glm<-RMSE(p_hat_glm,test_set$score)

#Creation of a table to store our results
rmse_table <- data_frame(Model= "GLM Model", RMSE= RMSE_glm)

rmse_table %>% knitr::kable()

```

The GLM Model shows some interesting results. The RMSE is low enough to accept the model. The most significant explanatory variables are log_pcgdp, life expectancy,freedom and corruption. 

##  2.Decision Tree Model

As a second model, we will try and fit the continuous variable decision tree model(DT)

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Fitting Decision Tree Regression to the dataset

#finding the best tune
set.seed(1)
train_rpart <- train(score ~ ., method = "rpart", tuneGrid = data.frame(cp = seq(0, 0.05, len = 25)), data = train_set)
ggplot(train_rpart)
train_rpart$bestTune


dt_method = rpart(formula = score ~ .,
                     data = train_set,
                     control = rpart.control(cp=0.0229,minsplit = 10))


# Predicting the Test set results
p_hat_dt = predict(dt_method, newdata = test_set)


RMSE_dt<-RMSE(p_hat_dt,test_set$score)

rmse_table <- bind_rows(rmse_table, data_frame(Model= "Decision Tree Model", RMSE=RMSE_dt))

rmse_table %>% knitr::kable()

# Plotting the tree
plot(dt_method, margin = 0.1)
text(dt_method, cex = 0.75)




```

The RMSE is low, though not as low as the GLM Model, is still low enough for us to accept the model. The decision tree tells us that happiness score of 7 is possible when Life Expectancy and Per-capita GDP are high. Similarly, lower Life Expectancy and lower Generosity scores can lead to a low Happiness Score of 4. 

An advantage of this model is the ease of interpretation and a couple of different scenarios about what variable mix can lead to a higher or lower happiness score.

##  3.Random Forest Model

Lastly we fit the Random Forest Modelto out training and testing datasets. 

```{r echo=FALSE, message=FALSE, warning=FALSE}

rf_method<-randomForest(score~.,data=train_set)

plot(rf_method)

p_hat_rf = predict(rf_method, newdata = test_set)

RMSE_rf<-RMSE(p_hat_rf,test_set$score)


rmse_table <- bind_rows(rmse_table, data_frame(Model= "Random Forest Model", RMSE=RMSE_rf))

rmse_table %>% knitr::kable()
```

We can infer from the figure above, that at about 200 trees, the error cannot be reduced any further.The RMSE obtained from fitting this model is low enough for the model to be acceptable. The Random Forest Model has a lower RMSE than the GLM Model.



#                   RESULTS AND CONCLUSIONS

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Final results table
rmse_table %>% knitr::kable()
```

In terms of the fitting a predictive model, all three models fitted very well with the training and testing sets with RMSE below the target of 0.8. 
The Random Forest Model did the best as it gave us the lowest RMSE.


#    Limitations and further exploration

The data set was quite small since there are only 138 rows, one for each country. Splitting the data into training and testing sets further reduced the accuracy of the models.

More explanatory variables, for example, Crime Statistics and Quality of Environment in a country, can be explored to explain happiness of people.

Also, more analysis can be done regarding inter-dependence among explanatory variables, as we found in the Correlation Plot.


#                              REFERENCES

Helliwell, John F., Richard Layard, Jeffrey Sachs, and Jan-Emmanuel De Neve, eds. 2020. World Happiness Report 2020. New York: Sustainable Development Solutions Network
 
The World Bank https://data.worldbank.org/indicator/SI.POV.GINI
 
CIA World Factbook

https://en.wikipedia.org/wiki/List_of_countries_by_income_equality


